Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-06-05 CGP Journal/main.tex
Added encode bar table
More in the behavior section

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 23 & 0.7174 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 20 & NA \\ \hline
\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 20 & 0.3125 \\ \hline
\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 22 & 0.3310 \\ \hline
\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 24 & 0.0248 \\ \hline
\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 17 & 0.0046 \\ \hline
\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 18 & 0 \\ \hline
\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline
	\end{tabular}
	\caption{Parity, 5.9956882435042246e-14}
	\label{tab:parity}
\end{table*}

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 49 & NA \\ \hline
\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 62 & 0.6918 \\ \hline
\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 64 & 0.8822 \\ \hline
\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 49 & 0.2525 \\ \hline
\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 69 & 0.0736 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 62 & 0.0211 \\ \hline
\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 32 & 0.0076 \\ \hline
\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 43 & 0.0085 \\ \hline
\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline
	\end{tabular}
	\caption{Encode 1.1926088284917749e-06}
	\label{tab:encode}
\end{table*}

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 114 & 0.5247 \\ \hline
\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 13,119 &   245 &  97 & 0.3610 \\ \hline
\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 11,991 &   211 &  87 & 0.8767 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 15,115 &   202 &  85 & NA \\ \hline
\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 14,590 &   203 &  85 & 0.7695 \\ \hline
\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 18,650 &   760 & 115 & 0.1868 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 29,432 & 4,983 & 125 & 0 \\ \hline
\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 125 & 0 \\ \hline
\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 &  97 & 0 \\ \hline
	\end{tabular}
	\caption{Decode 1.0182688124966139e-36}
	\label{tab:decode}
\end{table*}

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Reduced} & \textbf{p-value} \\ \hline
\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 484 & 0  \\ \hline
\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 85,970 & 3,117 & 489 & 0  \\ \hline
\emph{Single} & \emph{Dag} & 200  & NA & 239,844 & 78,739 & 114 & 74 & 0 \\ \hline
\emph{Single} & \emph{Reorder}    & 500   & NA & 243,664 & 91,935 & 266 & 131 & 0 \\ \hline
\emph{Skip} & \emph{Dag}          & 2,000 & 0.002 & 369,931 & 174,408 & 952 & 149 & 0.0100 \\ \hline
\emph{Accumulate} & \emph{Dag}    & 2,000 & 0.002  & 379,202 & 147,289 & 973 & 158 & 0.0053  \\ \hline
\emph{Single} & \emph{Normal}     & 5,000  & NA & 399,911 & 194,384 & 273 & 112 & 0.0047 \\ \hline
\emph{Accumulate} & \emph{Normal} & 2,000 & 0.002 & 483,938 & 227,686 & 168 & 87 & 0.45 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal} & 2,000 & 0.002 & 657,121 & 311,376 & 175 & 84 & NA \\ \hline

@@

\begin{table*}
	\centering
  \begin{tabular}{c|c|c|c|}
    \cline{2-4}
    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
    \thirdlabel{Skip} & \graphicthird{bar_multiply_skip_normal} &
                        \graphicthird{bar_multiply_skip_reorder} &
                        \graphicthird{bar_multiply_skip_dag}\\ \hline
    \thirdlabel{Accumulate} & \graphicthird{bar_multiply_accumulate_normal} &
                              \graphicthird{bar_multiply_accumulate_reorder} &
                              \graphicthird{bar_multiply_accumulate_dag}\\ \hline
    \thirdlabel{Single} & \graphicthird{bar_multiply_single_normal} &
                          \graphicthird{bar_multiply_single_reorder} &
                          \graphicthird{bar_multiply_single_dag}\\ \hline
	\end{tabular}
	\caption{Average node behavior for all nine algorithm combinations on the Multiply problem.
	         Excludes never active nodes.}
	\label{tab:behavior}
\end{table*}

Normal has the highest "used" bar unless its Single, likely because it also has the least active nodes

TODO High amount of TF.

TODO Breakdown of Active Nodes.

TODO Breakdown of Inactive Nodes.

TODO Look at how much of the genome is duplicated in both active and inactive

>>>>

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 23 & 0.7174 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 20 & NA \\ \hline
\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 20 & 0.3125 \\ \hline
\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 22 & 0.3310 \\ \hline
\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 24 & 0.0248 \\ \hline
\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 17 & 0.0046 \\ \hline
\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 18 & 0 \\ \hline
\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline
	\end{tabular}
	\caption{Parity, 5.9956882435042246e-14}
	\label{tab:parity}
\end{table*}

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 49 & NA \\ \hline
\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 62 & 0.6918 \\ \hline
\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 64 & 0.8822 \\ \hline
\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 49 & 0.2525 \\ \hline
\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 69 & 0.0736 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 62 & 0.0211 \\ \hline
\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 32 & 0.0076 \\ \hline
\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 43 & 0.0085 \\ \hline
\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline
	\end{tabular}
	\caption{Encode 1.1926088284917749e-06}
	\label{tab:encode}
\end{table*}

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Skip} & \emph{Reorder}       &  2,000 & 0.002 &  62,853 & 20,023 &   757 & 114 & 0.5247 \\ \hline
\emph{Single} & \emph{Reorder}     &    500 &    NA &  63,731 & 13,119 &   245 &  97 & 0.3610 \\ \hline
\emph{Accumulate} & \emph{Normal}  &  1,000 & 0.002 &  67,390 & 11,991 &   211 &  87 & 0.8767 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        &  1,000 & 0.002 &  68,231 & 15,115 &   202 &  85 & NA \\ \hline
\emph{Single} & \emph{Normal}      &  1,000 &    NA &  68,819 & 14,590 &   203 &  85 & 0.7695 \\ \hline
\emph{Accumulate} & \emph{Reorder} &  2,000 & 0.002 &  72,991 & 18,650 &   760 & 115 & 0.1868 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.001 & 132,821 & 29,432 & 4,983 & 125 & 0 \\ \hline
\emph{Accumulate} & \emph{Dag}     &  5,000 & 0.002 & 133,188 & 29,326 & 2,542 & 125 & 0 \\ \hline
\emph{Single} & \emph{Dag}         &    500 &    NA & 160,696 & 39,242 &   299 &  97 & 0 \\ \hline
	\end{tabular}
	\caption{Decode 1.0182688124966139e-36}
	\label{tab:decode}
\end{table*}

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Skip} & \emph{Reorder}      & 10,000 & 0.0005  & 196,116  & 60,326  & 3,175 & 484 & 0  \\ \hline
\emph{Accumulate} & \emph{Reorder}& 10,000 & 0.001 & 199,193 & 85,970 & 3,117 & 489 & 0  \\ \hline
\emph{Single} & \emph{Dag} & 200  & NA & 239,844 & 78,739 & 114 & 74 & 0 \\ \hline
\emph{Single} & \emph{Reorder}    & 500   & NA & 243,664 & 91,935 & 266 & 131 & 0 \\ \hline
\emph{Skip} & \emph{Dag}          & 2,000 & 0.002 & 369,931 & 174,408 & 952 & 149 & 0.0100 \\ \hline
\emph{Accumulate} & \emph{Dag}    & 2,000 & 0.002  & 379,202 & 147,289 & 973 & 158 & 0.0053  \\ \hline
\emph{Single} & \emph{Normal}     & 5,000  & NA & 399,911 & 194,384 & 273 & 112 & 0.0047 \\ \hline
\emph{Accumulate} & \emph{Normal} & 2,000 & 0.002 & 483,938 & 227,686 & 168 & 87 & 0.45 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal} & 2,000 & 0.002 & 657,121 & 311,376 & 175 & 84 & NA \\ \hline

@@

\begin{table*}
	\centering
  \begin{tabular}{c|c|c|c|}
    \cline{2-4}
    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
    \thirdlabel{Skip} & \graphicthird{bar_encode_skip_normal} &
                        \graphicthird{bar_encode_skip_reorder} &
                        \graphicthird{bar_encode_skip_dag}\\ \hline
    \thirdlabel{Accumulate} & \graphicthird{bar_encode_accumulate_normal} &
                              \graphicthird{bar_encode_accumulate_reorder} &
                              \graphicthird{bar_encode_accumulate_dag}\\ \hline
    \thirdlabel{Single} & \graphicthird{bar_encode_single_normal} &
                          \graphicthird{bar_encode_single_reorder} &
                          \graphicthird{bar_encode_single_dag}\\ \hline
	\end{tabular}
	\caption{Average node behavior for all nine algorithm combinations on the Encode problem.
	         Excludes never active nodes.}
	\label{tab:encode_behavior}
\end{table*}

\begin{table*}
	\centering
  \begin{tabular}{c|c|c|c|}
    \cline{2-4}
    & \textbf{\em Normal} & \textbf{\em Reorder} & \textbf{\em DAG} \\ \hline
    \thirdlabel{Skip} & \graphicthird{bar_multiply_skip_normal} &
                        \graphicthird{bar_multiply_skip_reorder} &
                        \graphicthird{bar_multiply_skip_dag}\\ \hline
    \thirdlabel{Accumulate} & \graphicthird{bar_multiply_accumulate_normal} &
                              \graphicthird{bar_multiply_accumulate_reorder} &
                              \graphicthird{bar_multiply_accumulate_dag}\\ \hline
    \thirdlabel{Single} & \graphicthird{bar_multiply_single_normal} &
                          \graphicthird{bar_multiply_single_reorder} &
                          \graphicthird{bar_multiply_single_dag}\\ \hline
	\end{tabular}
	\caption{Average node behavior for all nine algorithm combinations on the Multiply problem.
	         Excludes never active nodes.}
	\label{tab:multiply_behavior}
\end{table*}

In CGP it is common to report what fraction of the genome is active, as this represents
the easiest estimate of solution size.  By utilizing \emph{footprint} comparison
and the \Call{Simplify}{} algorithm, we can get a much better estimate of what
purpose each node in the genome serves.
Table~\ref{tab:parity}, Table~\ref{tab:encode}, Table~\ref{tab:decode},
and Table~\ref{tab:multiply} provide two column relevant to node behavior: Active
and Used.  The former gives the traditional measure of the median amount of
active nodes in the solution found.  The latter is the median number of nodes
still active after applying the \Call{Simplify}{} algorithm.

On all problems and for all algorithm combinations, the number of active nodes is
far smaller than the genome size, and the number of Used nodes
is far smaller than the number of Active nodes.  In some extreme cases less
than 1\% of active nodes in the evolved solution are actually necessary.  Only the
most extreme configurations have even 50\% of active nodes in use by the
simplified solution.

Table~\ref{tab:encode_behavior} and Table~\ref{tab:multiply_behavior} help illustrate
the behavioral breakdown of nodes for each algorithm combination on the Encode
and Multiply problems, respectively.  These two problems were chosen as they
represent different ends of the quality spectrum for \emph{Skip Normal} and
because together the are sufficiently representative of behaviors seen on the other
problems.

Each diagram describes node behavior as falling into eight distinct categories.
The first category are the constant nodes, both active and inactive.  These nodes
return the same value for all tested inputs.  The ``Used'' bar is the average
number of nodes which are part of the simplified solution.  The ``Useful'' bars
express how many nodes, either active or inactive, have identical \emph{footprint}
values to used nodes.  These nodes represent either independent methods of calculating
the same behavior or are effectively pass through nodes which mimic the behavior
of their inputs.  Active ``Intron'' nodes are those active nodes that do not
fall into either the constant or used categories, with inactive ``Intron'' nodes
being those inactive nodes that have identical \emph{footprint} values to active
intron nodes.  The final bar is ``Explore'', which marks how many nodes on
average produce a \emph{footprint} not part of any previous category.  Note that
all never active nodes are ignored by these plots, as their behavior is meaningless.

The first striking feature of these tables is how many nodes produce constant
outputs.  As the instruction set used in our genomes never require constant values
to produce any non constant behavior (TRUE NAND X $\Leftrightarrow$ X NAND X)
and most operations with constants result in pass through nodes
(FALSE OR X $\Leftrightarrow$ X), we would not expect evolution to proliferate
these constant nodes.  If these constants are in fact of no use, future work
could likely improve CGP's performance by reducing their prevalence.  Yet
it is possible they serve an evolutionary function by allowing mutation more
flexibility.  For instance, pass through nodes may be allowing mutation to create
adjustments not easily done if this behavior was forbidden.

As with previous analysis, \emph{Skip} and \emph{Accumulate} have nearly identical
plots, regardless of ordering method and problem.  \emph{Single} again has
a significantly different plot when paired with \emph{Reorder} or \emph{DAG},
with some hits of a change on Encode with \emph{Normal}.  These combinations
were also the most likely to have significantly smaller genome sizes.  From the
results tables we see that the number of used nodes on each problem is relatively
constant across the combinations even though the number of active nodes and the
genome sizes changed significantly.  As such, the proportion of nodes in the
used category will vary as these other features change.

From Table~\ref{tab:encode_behavior} and Table~\ref{tab:multiply_behavior}, having
the plurality of active nodes in the used category appears to correlate with
increase evaluations to success.  Notice that in all \emph{Single} plots the
tallest bar for active nodes is ``Used,'' and the only non \emph{Single} plots
where this is true are \emph{Normal} on Multiply.  On Encode, \emph{Single}
runs had the highest median evaluations to success, with \emph{Normal} runs
doing similarly on Multiply.  This appears to imply that having duplication
of node behavior in the active nodes, and not just inactive nodes, is beneficial
to search.  If this is true, it may help explain CGP uses such large genomes,
as \emph{Normal} is only able to achieve enough active nodes to allow duplication
when the genome size significantly exceeds the actual solution size.  Also
the success of \emph{Reorder} and \emph{DAG} on Multiply may also be partially be attributable
to this ability to increase the number of active nodes.

One significant difference in behavior between the two problems is proportion of
explore nodes.  A potential explanation is that on 16-4-bit Encode there are only
$2^{16} = 65,536$ \emph{footprints} while on 3-bit Multiply there are $2^{64} = 1.845\cdot 10^{19}$
\emph{footprints}.  Combined with the fact that not all \emph{footprints} are equally
likely to be produced, the difference may just be that on Encode nodes are
that much more likely to recreate existing behavior.  Another possibility is that
CGP increases the diversity of behavior of inactive nodes over evolutionary time.
As such, the fact that Multiply required approximately an order of magnitude more
evaluations to solve would lead to the discrepancy.
